`infercnv` results exploration for `r params$sample_id`

### 2024-10-21 analyses/cell-type-wilms-tumor-06/notebook_template/06_cnv_infercnv_exploration.Rmd
improve PR#828 and try inter-patient normal reference

---
title: "`infercnv` results exploration for `r params$sample_id`"
author: "Maud PLASCHKA"
date: "`r Sys.Date()`"
params:
  scpca_project_id: "SCPCP000006"
  sample_id: "SCPCS000179"
  seed: 12345
output: 
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

@@

`infercnv` was run using the `06_inferCNV.R` script with and without a normal reference. We tested the impact of the sub-selection of normal cells using either immune, and/or endothelial cells as healthy reference.
In addition, we are exploring the use of the [HMM based CNV Prediction Methods](https://github.com/broadinstitute/infercnv/wiki/inferCNV-HMM-based-CNV-Prediction-Methods).

@@

library(infercnv)
library(SCpubr)
library(ggplot2)
library(patchwork)
library(Seurat)

@@

@@

### Try to calculate  single CNV score and see if it can be use to define cells with CNV

@@

This unique CNV score do not look so promising. 
We might have to select chromosomes we would like to look at, i.e. the one relevant for Wilms tumor.

### Explore infercnv results generated with immune and endothelial cells as reference, using a HMM i3 model

@@

srat <- readRDS(file.path(module_base, "results", params$sample_id, glue::glue("06_infercnv_HMM-i3_", params$sample_id, "_reference-both.rds")))

@@

@@

### Explore infercnv results generated with immune and endothelial cells as reference, using a HMM i6 model

srat <- readRDS(file.path(module_base, "results", params$sample_id, glue::glue("06_infercnv_HMM-i6_", params$sample_id, "_reference-both.rds")))

@@

@@

- Providing no reference is not a good option, as we think that most of the cells are cancer cells with few CNV. 
Without healthy reference, `infercnv` take the mean of expression across the cells as a reference. 
In case of events that are well shared by _a majority_ of the cells, they will be smoothed.

- For some CNV and specific chromosome, the choice a the reference do not affect the result: ch18, ch17.
- Some CNV values differs depending on the choice of the reference, for chr5 for example.

This is an issue if we want to have the exact CNV pattern.

There are three samples with a very low (<5) number of immune + endothelial cells: SCPCS000197 (4 endothelial), SCPCS000190 (1 immune), SCPCS000177	(2 immune) for which it might be problematic to run `infercnv`.

I am not sure if some endothelial cells might be missclassified? 
To be careful when we look at % of alterations within a compartment, as there are sometimes very few immune and/or endothelial cells, some % can be easily high without being meaningful.

HMM-i3 model looks "cleaner" than the HMM-i6 model, with less CNV found in immune and endothelial cells.

>>>>

---
title: "`infercnv` results exploration for `r params$sample_id`"
author: "Maud PLASCHKA"
date: "`r Sys.Date()`"
params:
  scpca_project_id: "SCPCP000006"
  sample_id: "SCPCS000184"
  seed: 12345
output: 
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

@@

`infercnv` was run using the `06_inferCNV.R` script with and without a normal reference, from the same patient or from an inter-patient pull of normal cells. We tested the impact of the sub-selection of normal cells using either immune, and/or endothelial cells as healthy reference.
In addition, we are exploring the use of the [HMM based CNV Prediction Methods](https://github.com/broadinstitute/infercnv/wiki/inferCNV-HMM-based-CNV-Prediction-Methods).

@@

library(infercnv)
library(SCpubr)
library(ggplot2)
library(patchwork)
library(Seurat)
```

@@

#### With immune and endothelium cells from all upfront resection samples spiked-in as reference

![](`r file.path(result_dir, "reference-pull_HMM-i3" , glue::glue("06_infercnv_", params$sample_id, "_", "reference-pull_HMM-i3", "_heatmap.png"))`)

@@

```
These heatmaps emphasize the importance of the selection of normal cells prior the inference of CNV.
The normal reference should contain as much cell types as possible, in order to minimize false positive CNV.
In our case, we should take immune and entodethial cells when possible.

Of note: By default if no reference is provided, `infercnv` take the mean of expression as normal reference. 
The risk is that the main cell population (in our case the fetal nephron compartment) might be mistaken as the normal baseline.

### Summary CNV score

We want to calculate a single CNV score and asess if/how it can be use to define cells with CNV versus stable/normal cells. 
We defined the score as discribed in the [biostar discussion](https://www.biostars.org/p/9573777/).

We would expect:

- immune and endothelial cells to have a low CNV score

- fetal nephron and stroma cells to have low to high CNV score.

Ideally, the CNV score distribution should show two peaks, one for the normal cells with no CNV, one for cancer cells with more genomic rearrangements.

@@

This unique CNV score does not look promising. 
We will not use this score in future notebooks.

We might have to select chromosomes we would like to look at, i.e. the one relevant for Wilms tumor (1p, 1q, chr17).

### HMM-i3 inference prediction with both immune and endothelium cells as reference

We then explore infercnv results generated with immune and endothelial cells as reference, using a [HMM-i3 prediction models](https://github.com/broadinstitute/infercnv/wiki/infercnv-i3-HMM-type).

@@

srat_i3 <- readRDS(file.path(module_base, "results", params$sample_id, glue::glue("06_infercnv_HMM-i3_", params$sample_id, "_reference-both.rds")))

@@

Immune and endothelial cells shouldn't show any CNV.

The stroma and fetal nephron compartments may contain normal cells with no CNV or cancer cells with low to high CNV profiles.

visualize_metadata(srat_i3, "fetal_kidney_predicted.compartment", "seurat_clusters")

This should allow a quick visual check that immune and endothelial cells do not expressed any CNV. 
If not, we should check if genes driving the CNV score are not key immune or endothelium genes.

The `umap` reduction visualization might allow the identification of cluster(s) of normal cells with no CNV.

```{r fig.width=10, fig.height=5, out.width='45%', out.height='45%'}
for(i in 1:22){
  print(visualize_feature(srat_i3, features=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment"))
}
```

####  CNV distribution in the Wilms tumor compartments

```{r fig.width=5, fig.height=5, out.width='40%', out.height='40%', warning=FALSE}
for(i in 1:22){
  print(visualize_density(srat_i3, features=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment"))
}
```

#### DotPlot

The `Dotplot` representation summarizes the percentage of cells in each compartment with cnv in each of the 22 chromosomes.

The HMM-i3 prediction model with intra-patient reference (composed of immune and endothelium cells) perform really well in identifying Wilms tumor relevant CNV.
No false positive CNV are observed in the immune and endothelium compartment.

```{r fig.width=15, fig.height=5, out.width='100%'}

DotPlot(srat_i3, features =  colnames(srat_i3@meta.data)[grepl("has_cnv_chr", colnames(srat_i3@meta.data))], group.by = "fetal_kidney_predicted.compartment", assay = "SCT") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

#### CNV score

Here, we try to calculate a single CNV score and assess its potential in iditenfying cells with CNV versus normal cells without CNV.

We simply checked for each chromosome if the cell `has_cnv_chr`. 
Would the cell have one or more chromosome with CNV, the global `has_cnv_score` will be TRUE.
Else, the cell will have a `has_cnv_score` set to FALSE.

```{r fig.width=20, fig.height=5, out.width='100%', out.height='100%', warning=FALSE}
meta <- srat_i3@meta.data

meta$has_cnv_score <- rowSums(meta[,grepl("has_cnv_chr", colnames(meta))])
meta$has_cnv_score[meta$has_cnv_score > 1] <- "CNV"

# we know that some immune cells might have false positive CNV 
meta$has_cnv_score[meta$has_cnv_score %in% c(0,1)] <- "no CNV"

srat_i3 <- AddMetaData(srat_i3, meta$has_cnv_score, col.name =  "has_cnv_score")

We identified clusters with no/low CNV while the majority of the `Seurat` cluster present CNV on more than one chromosome.

visualize_metadata(srat_i3, "has_cnv_score", "seurat_clusters")

Immune and endothelial cells have a majority of cells with no/low CNV profile, which confirm their annotation as normal cells. 
The fetal nephron and stroma copartments are mainly composed of cancer cells with more than one chromosome affected by a CNV.

visualize_metadata(srat_i3, "has_cnv_score", "fetal_kidney_predicted.compartment")

### HMM-i6 prediction of CNV with intra-patient normal reference (immune and endothelial)

We then explore `infercnv` results generated with immune and endothelial cells as reference, using a [HMM-i6 model](https://github.com/broadinstitute/infercnv/wiki/infercnv-i6-HMM-type).

srat_i6 <- readRDS(file.path(module_base, "results", params$sample_id, glue::glue("06_infercnv_HMM-i6_", params$sample_id, "_reference-both.rds")))

```{r fig.width=10, fig.height=5, out.width='45%', out.height='45%'}
for(i in 1:22){
  print(visualize_feature(srat_i6, features=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment"))
}
```

```{r fig.width=5, fig.height=5, out.width='40%', out.height='40%', warning=FALSE}
for(i in 1:22){
  print(visualize_density(srat_i6, features=glue::glue("proportion_cnv_chr", i), group.by = "fetal_kidney_predicted.compartment"))
}
```

#### DotPlot

The `Dotplot` representation summarizes the percentage of cells in each compartment with cnv in each of the 22 chromosomes.

The HMM-i3 prediction model with intra-patient reference (composed of immune and endothelium cells) perform really well in identifying Wilms tumor relevant CNV.
No false positive CNV are observed in the immune and endothelium compartment.

```{r fig.width=15, fig.height=5, out.width='100%'}

DotPlot(srat_i6, features =  colnames(srat_i6@meta.data)[grepl("has_cnv_chr", colnames(srat_i6@meta.data))], group.by = "fetal_kidney_predicted.compartment", assay = "SCT") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

#### CNV score

Finally, we try to calculate  single CNV score and assess its potential in iditenfying cells with CNV versus normal cells without CNV.

We simply checked for each chromosome if the cell `has_cnv_chr`. 
Would the cell have one or more chromosome with CNV, the global `has_cnv_score` will be TRUE.
Else, the cell will have a `has_cnv_score` set to FALSE.

```{r fig.width=20, fig.height=5, out.width='100%', out.height='100%', warning=FALSE}
meta <- srat_i6@meta.data

meta$has_cnv_score <- rowSums(meta[,grepl("has_cnv_chr", colnames(meta))])
meta$has_cnv_score[meta$has_cnv_score > 1] <- "CNV"

# we know that some immune cells might have false positive CNV 
meta$has_cnv_score[meta$has_cnv_score %in% c(0,1)] <- "no CNV"

srat_i6 <- AddMetaData(srat_i6, meta$has_cnv_score, col.name =  "has_cnv_score")

visualize_metadata(srat_i6, "has_cnv_score", "seurat_clusters")

visualize_metadata(srat_i6, "has_cnv_score", "fetal_kidney_predicted.compartment")

### HMM-i3 CNV prediction model with an inter-patient reference of immune cells (endothelial and immune)

We lastly explore `infercnv` results generated with a pull of immune and endothelial cells from different Wilms tumor samples as reference, using a HMM-i3 prediction model.

The selection of normal cells to be spiked-in as a normal reference prior running `infercnv` has been performed in `06a_build-normal-cell_reference.R.
We selected samples that haven't been pre-treated with chemotherapies. Indeed, even if normal cells shouldn't be affected by chemotherapy in terms of CNV, we are not sure how chemotherapies can affect the CNV profile of normal cells.
We thus decided to only take endothelial and immune cells from non-treated samples to build the reference of normal cells.

We load the `Seurat` object generated in `06_infercnv.R` and transfer the `infercnv` metadata to the main `Seurat` object of sample `r params$sample_id` saved in `r glue::glue("02b-fetal_kidney_label-transfer_", params$sample_id, ".Rds")`.

Such a transfer is required as the merged `Seurat` object + spike-in cells does not contained the `umap` reduction.

```{r}
seurat_full <- readRDS(file.path(module_base, "results", params$sample_id, glue::glue("06_infercnv_HMM-i3_", params$sample_id, "_reference-pull.rds")))
# subset non spike-in cells
keep_cells <- colnames(seurat_full)[!grepl("spike", colnames(seurat_full))]
seurat <- subset(seurat_full, cells = keep_cells)

meta <- seurat@meta.data

# merge the new info into the srat metadata
srat <- AddMetaData(srat, meta[,grepl("_cnv_chr", colnames(meta))], col.name = colnames(meta[grepl("_cnv_chr", colnames(meta))]))
```

@@

#### DotPlot

The `Dotplot` representation summarizes the percentage of cells in each compartment with cnv in each of the 22 chromosomes.

The HMM-i3 prediction model with inter-patient reference (composed of immune and endothelium cells) perform quite well in identifying Wilms tumor relevant CNV.
However, sample variation induced false positive CNV such as CNV on chromosome 5 and 13 observed in the endothelium compartment.

```{r fig.width=15, fig.height=5, out.width='100%'}

DotPlot(srat, features =  colnames(srat@meta.data)[grepl("proportion_cnv_chr", colnames(srat@meta.data))], group.by = "fetal_kidney_predicted.compartment", assay = "SCT") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

@@

#### CNV score

Finally, we try to calculate  single CNV score and assess its potential in iditenfying cells with CNV versus normal cells without CNV.

We simply checked for each chromosome if the cell `has_cnv_chr`. 
Would the cell have three or more chromosome with CNV, the global `has_cnv_score` will be TRUE.
Else, the cell will have a `has_cnv_score` set to FALSE.

Of note, as we know that the use of a ~universal inter-patient reference of normal cells can induce false positive CNV in some chromosomes, we decided to allow up to 2 CNV in normal cells.

```{r fig.width=20, fig.height=5, out.width='100%', out.height='100%', warning=FALSE}
meta <- srat@meta.data
meta$has_cnv_score <- rowSums(meta[,grepl("has_cnv_chr", colnames(meta))])
meta$has_cnv_score[meta$has_cnv_score > 2] <- "CNV"

# we know that some immune cells might have false positive CNV 
meta$has_cnv_score[meta$has_cnv_score %in% c(0,1, 2)] <- "no CNV"

srat <- AddMetaData(srat, meta$has_cnv_score, col.name =  "has_cnv_score")

visualize_metadata(srat, "has_cnv_score", "seurat_clusters")

visualize_metadata(srat, "has_cnv_score", "fetal_kidney_predicted.compartment")

## Comparisons of inter- and intra-patient global cnv score with HMM prediction model

We compare here the binary CNV scores calculated with the three HMM prediction models:

- HMM-i3 with inter-patient endothelial and immunce cells as reference

- HMM-i3 with intra-patient endothelial and immunce cells as reference

- HMM-i6 with intra-patient endothelial and immunce cells as reference

p1 <- SCpubr::do_DimPlot(srat, "has_cnv_score", plot.title = "inter-patient reference, HMM-i3", reduction = "umap")

p2 <- SCpubr::do_DimPlot(srat_i3, "has_cnv_score", plot.title = "intra-patient reference, HMM-i3", reduction = "umap")

p3 <- SCpubr::do_DimPlot(srat_i6, "has_cnv_score", plot.title = "intra-patient reference, HMM-i6", reduction = "umap")

p1 + p2 + p3 + plot_layout(ncol = 3)

@@

- We should run `infercnv` with the most diverse normal reference as possible, including different cell types.
Providing no reference is not a good option, as we think that most of the cells are cancer cells with few CNV. 
In our case, we advise taking at least immune and endothelial cells as normal reference.

- The HMM prediction models help exploring the `infercnv` results. 
In this notebook, we have compared three HMM prediction models:

   + HMM-i3 with inter-patient endothelial and immunce cells as reference

   + HMM-i3 with intra-patient endothelial and immunce cells as reference

   + HMM-i6 with intra-patient endothelial and immunce cells as reference

- Globally, the three scores seems to drive similar conclusions, with the majority of fetal nephron and stroma cells being cancer cells, at least in the sample selected.

   + The HMM-i3 model with inter-patient endothelial and immunce cells as reference has the advantage to be usable for all Wilms tumor samples, including the ones with a very low number of immune and/or endothelial cells.

   + The HMM-i3 model with intra-patient endothelial and immunce cells as reference seems to be the cleaner, ~fast to run (10 minutes per samples) and is more precise than the HMM-i3 with the inter-patient reference.

   + The HMM-i6 model with intra-patient endothelial and immunce cells as reference is very slow (~2 hours per sample) and couldn't be used for the entire cohorte.
It is more noisy than the i3 version. However, it could have the potential to detect cancer cells with very low CNV profile.

- Surprisingly, running `infercnv` with emdothelial and immune cells from (i) the same patient or (ii) a set of Wilms tumor patients do not seem to affect drastically the results. 
Some false positive CNV might occur in every patient due to the inter-patient variability. 
By comparing the results in conditions (i) and (ii), we should be able to understand which false positive are recurrent and do not take them into account.
